---
title: "anansi example: top networks per reticulation count (d10k_x100), outgroup-snap view"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)
library(ggplot2)

# Use the installed package if available, else load the source tree. This makes
# the notebook work both as a dev example (knit from the repo) and standalone.
if (requireNamespace("anansi", quietly = TRUE)) {
  library(anansi)
} else {
  devtools::load_all(if (file.exists("DESCRIPTION")) "." else "..")
}

# Resolve the data path whether knit from the repo root or from notebooks/.
data_path <- if (file.exists("data/d10k_x100_results.csv")) {
  "data/d10k_x100_results.csv"
} else {
  "../data/d10k_x100_results.csv"
}
```

## What this notebook does

`data/d10k_x100_results.csv` holds inferred phylogenetic networks for two
reference rootings of the same 15 lizard taxa --- a *Lacerta agilis* reference
(`dataset == "lac"`) and a *Zootoca vivipara* reference (`dataset == "zoo"`) ---
each with the **top 10 networks per reticulation count** (1, 2, and 3
reticulations) ranked by `log_probability`.

For each reference we:

1. keep the **top 3 networks per reticulation count** (9 networks per reference), then
2. draw an anansi `densinet()` **consensus overlay** that pins the outgroup
   (*Zootoca vivipara*, *Lacerta agilis*) to the top and snaps the shared tip
   order to the consensus topology so the consensus backbone has no crossings
   (the same style as `figures/11_lacerta_snap.png`).

The consensus backbone is the majority-rule tree of the networks' backbones,
shaded grey&rarr;black by clade support; the reticulation events frequent across
the selected networks are drawn as arrows.

## Load the networks

```{r load}
all <- read_networks_csv(data_path)
all
table(dataset = all$meta$dataset, reticulations = all$meta$reticulations)
```

## Select the top 3 networks per reticulation count

`top_networks()` ranks across a whole set; here we want the best 3 *within each
reticulation count*, separately per reference. A small grouped selection on the
metadata does that, then we subset the netset with `[`.

```{r select}
top_per_reticulation <- function(ns, dataset, n = 3) {
  sub <- ns[which(ns$meta$dataset == dataset)]
  m <- sub$meta
  idx <- unlist(
    lapply(split(seq_len(nrow(m)), m$reticulations), function(ii) {
      ii[order(m$log_probability[ii], decreasing = TRUE)][seq_len(min(n, length(ii)))]
    }),
    use.names = FALSE
  )
  sub[sort(idx)]
}

lac <- top_per_reticulation(all, "lac", n = 3)
zoo <- top_per_reticulation(all, "zoo", n = 3)

lac$meta[, c("dataset", "reticulations", "rank", "log_probability")]
```

## Outgroup-snap consensus plots

```{r plot-params}
outgroup <- c("Zootoca_vivipara", "Lacerta_agilis")
```

### Lacerta reference

```{r plot-lac}
densinet(
  lac,
  method            = "mds",
  outgroup          = outgroup,
  outgroup_position = "top",
  snap_to_consensus = TRUE,
  title = "d10k_x100 (lac ref): top 3 networks per reticulation count"
)
```

### Zootoca reference

```{r plot-zoo}
densinet(
  zoo,
  method            = "mds",
  outgroup          = outgroup,
  outgroup_position = "top",
  snap_to_consensus = TRUE,
  title = "d10k_x100 (zoo ref): top 3 networks per reticulation count"
)
```

## Notes

- `outgroup` / `outgroup_position = "top"` pin *Zootoca* and *Lacerta* to the
  top of each figure; `snap_to_consensus = TRUE` rotates the shared tip order to
  the consensus topology so the (black) consensus backbone has no crossings.
- Because we overlay the top 3 networks for **each** of 1--3 reticulations, the
  reticulation cloud (and the consensus reticulation arrows) reflects events
  across all those models at once. Tune `consensus_ret_min` to keep only the
  most frequent events, or set `reticulation_style = "hybrid"` for the
  direction-agnostic two-edge layout.
- To anchor reticulations further along their branches, raise `ret_edge_frac`
  (default `0.1`).
```
