---
title: "anansi example: top networks per reticulation count (d10k_x100), outgroup-snap view"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6, dpi = 300
)
library(ggplot2)

# Use the installed package if available, else load the source tree. This makes
# the notebook work both as a dev example (knit from the repo) and standalone.
if (requireNamespace("anansi", quietly = TRUE)) {
  library(anansi)
} else {
  devtools::load_all(if (file.exists("DESCRIPTION")) "." else "..")
}

# Resolve the data path whether knit from the repo root or from notebooks/.
data_path <- if (file.exists("data/d10k_x100_results.csv")) {
  "data/d10k_x100_results.csv"
} else {
  "../data/d10k_x100_results.csv"
}
```

## What this notebook does

`data/d10k_x100_results.csv` holds inferred phylogenetic networks for two
reference rootings of the same 15 lizard taxa --- a *Lacerta agilis* reference
(`dataset == "lac"`) and a *Zootoca vivipara* reference (`dataset == "zoo"`) ---
each with the **top 10 networks per reticulation count** (1, 2, and 3
reticulations) ranked by `log_probability`.

We keep the **top 3 networks per reticulation count from each reference** and
**join both references into one set** (18 networks), then draw a single anansi
`densinet()` **consensus overlay** that pins the outgroup (*Zootoca vivipara*,
*Lacerta agilis*) to the top and snaps the shared tip order to the consensus
topology so the consensus backbone has no crossings (the same style as
`figures/11_lacerta_snap.png`).

The consensus backbone is the majority-rule tree of the networks' backbones,
shaded grey&rarr;black by clade support; the reticulation events frequent across
the selected networks are drawn as arrows.

## Load the networks

```{r load}
all <- read_networks_csv(data_path)
all
table(dataset = all$meta$dataset, reticulations = all$meta$reticulations)
```

## Select the top 3 per reticulation count, both references combined

`top_networks()` ranks across a whole set; here we want the best 3 *within each
reticulation count of each reference*, then everything pooled into one netset.
A grouped selection on the metadata gives the row indices, and subsetting the
netset with `[` returns a single combined `anansi_netset` (with metadata kept in
sync).

```{r select}
top_per_reticulation <- function(ns, n = 3, datasets = unique(ns$meta$dataset)) {
  m <- ns$meta
  keep <- which(m$dataset %in% datasets)
  grp <- interaction(m$dataset[keep], m$reticulations[keep], drop = TRUE)
  idx <- unlist(
    lapply(split(keep, grp), function(ii) {
      ii[order(m$log_probability[ii], decreasing = TRUE)][seq_len(min(n, length(ii)))]
    }),
    use.names = FALSE
  )
  ns[sort(idx)]
}

combined <- top_per_reticulation(all, n = 3)   # 3 per (reference x reticulation count)
combined
table(dataset = combined$meta$dataset, reticulations = combined$meta$reticulations)
```

## Outgroup-snap consensus plot (both references combined)

```{r plot-combined}
densinet(
  combined,
  method            = "mds",
  outgroup          = c("Zootoca_vivipara", "Lacerta_agilis"),
  outgroup_position = "top",
  ret_edge_frac     = 0.75,
  consensus_ret_min = 0.25,
  snap_to_consensus = TRUE,
  title = "d10k_x100: top 3 per reticulation count (both references combined)"
)
```

## Notes

- `combined` pools the *Lacerta*- and *Zootoca*-reference selections into one
  `anansi_netset`; the consensus backbone and reticulation cloud therefore
  summarize **agreement across both references and all reticulation counts** at
  once. To inspect a single reference, pass `datasets = "lac"` (or `"zoo"`) to
  the helper.
- `outgroup` / `outgroup_position = "top"` pin *Zootoca* and *Lacerta* to the
  top; `snap_to_consensus = TRUE` rotates the shared tip order to the consensus
  topology so the (black) consensus backbone has no crossings.
- Tune `consensus_ret_min` to keep only the most frequent reticulation events,
  set `reticulation_style = "hybrid"` for the direction-agnostic two-edge
  layout, or raise `ret_edge_frac` (default `0.1`) to anchor reticulations
  further along their branches.
```
